---
title: "ch10-Data-Viz-R"
author: "jenngeorgevich"
date: "9/17/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library('ggplot2')
```
## Scatter Plots 
- 

```{r faithful}
plot(faithful$waiting,
faithful$eruptions,
xlab = "waiting time (min)",
ylab = "eruption time (min)")
```

- args pch, col, cex modify shape, color, size 

```{r faithful}
plot(faithful$waiting,
faithful$eruptions,
pch = 17,
col = 2,
cex = 1.2,
xlab = "waiting times (min)",
ylab = "eruption time (min)")
```
 - distinguish between different values with different symbols like 'pch = mtcars$am'
 
 - qplot version: 
 
```{r faithful} 
qplot(x = waiting,
y = eruptions,
data = faithful,
main = "Waiting times (sec) vs. eruptions (min)")
```

- 'size' determines the size of each marker 

```{r mtcars}
qplot(x = wt,
y = mpg,
data = mtcars,
size = cyl,
main = "MPG vs. weight (x1000 lbs) by cylinder")
```

### Adding a smooth, curved line
- use a kernel function to get the weighted avg of the original data 
- modify the 'span' argument (similar to 'adjust' for smoothed histograms) to adjust 'h'

```{r mtcars}
qplot(disp,
mpg,
data = mtcars,
main = "MPG vs Eng. Displacement") +
stat_smooth(method = "loess",
method.args = list(degree = 0),
span = 0.2,
se = FALSE)
```
- the above plot's `span' is less than the below plot's. increasing span results in a less wiggly curve

```{r mtcars}
qplot(disp,
mpg,
data = mtcars,
main = "MPG vs Eng. Displacement") +
stat_smooth(method = "loess",
method.args = list(degree = 0),
span = 1,
se = FALSE)
```

### using qplots 'facets' to compare multiple graphs 
  - for side-by-side panels, use `facet = .~amf`
```{r mtcars}
# add new dataframe columns with more appropriate names for
# better axes labeling in future graphs
mtcars$amf[mtcars$am==0] = 'automatic'
mtcars$amf[mtcars$am==1] = 'manual'
mtcars$vsf[mtcars$vs==0] = 'flat'
mtcars$vsf[mtcars$vs==1] = 'V-shape'
qplot(x = wt,
y = mpg,
facets = .~amf,
data = mtcars,
main = "MPG vs. weight by transmission")
```
- for graphs stacked on top of each other, use `facet = vsf~.`
```{r mtcars}
mtcars$amf[mtcars$am==0] = 'automatic'
mtcars$amf[mtcars$am==1] = 'manual'
mtcars$vsf[mtcars$vs==0] = 'flat'
mtcars$vsf[mtcars$vs==1] = 'V-shape'
qplot(x = wt,
y = mpg,
facets = vsf~.,
data = mtcars,
main = "MPG vs. weight by engine") +
stat_smooth(se = FALSE)
```
- multi-row and multi-colum: `facets = vsf~amf`
```{r mtcars}
mtcars$amf[mtcars$am==0] = 'automatic'
mtcars$amf[mtcars$am==1] = 'manual'
mtcars$vsf[mtcars$vs==0] = 'flat'
mtcars$vsf[mtcars$vs==1] = 'V-shape'
qplot(x = wt,
y = mpg,
data = mtcars,
facets = vsf~amf,
main = "MPG vs. weight by transmission and engine")
```

### using plot or ggpairs to compare multiple graphs 
- `plot` receives an entire df as an argument 

```{r mtcars}
# create a new dataframe with three columns: cty, hwy, and disp
DF = mpg[, c("cty", "hwy", "displ")]
plot(DF, main = "City MPG vs. Highway MPG vs. Engine Volume")
```

```{r mtcars}
library(GGally)
DF = mpg[, c("cty", "hwy", "displ")]
ggpairs(DF)
```

## Contour Plots 
- graph a 2d function f(x, y) for different output values 
- create a df with columns for the x, y, and z values where the x and y columns are all possible combinations of the two coords over a certain grid. 

```{r plot}
x_grid = seq(-1, 1, length.out = 100)
y_grid = x_grid
# create a dataframe containing all possible combinations of x,y
R = expand.grid(x_grid, y_grid)
# number of rows is 100 x 100 - one for each combination
dim(R)
## [1] 10000 2
# modify column names for clear axes labeling
names(R) = c('x', 'y')
R$z = R$x^2 + R$y^2
head(R)
## x y z
## 1 -1.0000000 -1 2.000000
## 2 -0.9797980 -1 1.960004
## 3 -0.9595960 -1 1.920824
## 4 -0.9393939 -1 1.882461
## 5 -0.9191919 -1 1.844914
## 6 -0.8989899 -1 1.808183
ggplot(R, aes(x = x,y = y, z = z)) + stat_contour()
```


## Quantiles and Box Plots 
- 

