---
title: "ch10-Data-Viz-R"
author: "jenngeorgevich"
date: "9/17/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library('ggplot2')
```
## Scatter Plots 
- 

```{r faithful}
plot(faithful$waiting,
faithful$eruptions,
xlab = "waiting time (min)",
ylab = "eruption time (min)")
```

- args pch, col, cex modify shape, color, size 

```{r faithful}
plot(faithful$waiting,
faithful$eruptions,
pch = 17,
col = 2,
cex = 1.2,
xlab = "waiting times (min)",
ylab = "eruption time (min)")
```
 - distinguish between different values with different symbols like 'pch = mtcars$am'
 
 - qplot version: 
 
```{r faithful} 
qplot(x = waiting,
y = eruptions,
data = faithful,
main = "Waiting times (sec) vs. eruptions (min)")
```

- 'size' determines the size of each marker 

```{r mtcars}
qplot(x = wt,
y = mpg,
data = mtcars,
size = cyl,
main = "MPG vs. weight (x1000 lbs) by cylinder")
```

### Adding a smooth, curved line
- use a kernel function to get the weighted avg of the original data 
- modify the 'span' argument (similar to 'adjust' for smoothed histograms) to adjust 'h'

```{r mtcars}
qplot(disp,
mpg,
data = mtcars,
main = "MPG vs Eng. Displacement") +
stat_smooth(method = "loess",
method.args = list(degree = 0),
span = 0.2,
se = FALSE)
```
- the above plot's `span' is less than the below plot's. increasing span results in a less wiggly curve

```{r mtcars}
qplot(disp,
mpg,
data = mtcars,
main = "MPG vs Eng. Displacement") +
stat_smooth(method = "loess",
method.args = list(degree = 0),
span = 1,
se = FALSE)
```

### using qplots 'facets' to compare multiple graphs 
  - for side-by-side panels, use `facet = .~amf`
```{r mtcars}
# add new dataframe columns with more appropriate names for
# better axes labeling in future graphs
mtcars$amf[mtcars$am==0] = 'automatic'
mtcars$amf[mtcars$am==1] = 'manual'
mtcars$vsf[mtcars$vs==0] = 'flat'
mtcars$vsf[mtcars$vs==1] = 'V-shape'
qplot(x = wt,
y = mpg,
facets = .~amf,
data = mtcars,
main = "MPG vs. weight by transmission")
```
- for graphs stacked on top of each other, use `facet = vsf~.`
```{r mtcars}
mtcars$amf[mtcars$am==0] = 'automatic'
mtcars$amf[mtcars$am==1] = 'manual'
mtcars$vsf[mtcars$vs==0] = 'flat'
mtcars$vsf[mtcars$vs==1] = 'V-shape'
qplot(x = wt,
y = mpg,
facets = vsf~.,
data = mtcars,
main = "MPG vs. weight by engine") +
stat_smooth(se = FALSE)
```
- multi-row and multi-colum: `facets = vsf~amf`
```{r mtcars}
mtcars$amf[mtcars$am==0] = 'automatic'
mtcars$amf[mtcars$am==1] = 'manual'
mtcars$vsf[mtcars$vs==0] = 'flat'
mtcars$vsf[mtcars$vs==1] = 'V-shape'
qplot(x = wt,
y = mpg,
data = mtcars,
facets = vsf~amf,
main = "MPG vs. weight by transmission and engine")
```

### using plot or ggpairs to compare multiple graphs 
- `plot` receives an entire df as an argument 

```{r mtcars}
# create a new dataframe with three columns: cty, hwy, and disp
DF = mpg[, c("cty", "hwy", "displ")]
plot(DF, main = "City MPG vs. Highway MPG vs. Engine Volume")
```

```{r mtcars}
library(GGally)
DF = mpg[, c("cty", "hwy", "displ")]
ggpairs(DF)
```

## Contour Plots 
- graph a 2d function f(x, y) for different output values 
- create a df with columns for the x, y, and z values where the x and y columns are all possible combinations of the two coords over a certain grid. 

```{r plot}
x_grid = seq(-1, 1, length.out = 100)
y_grid = x_grid
# create a dataframe containing all possible combinations of x,y
R = expand.grid(x_grid, y_grid)
# number of rows is 100 x 100 - one for each combination
dim(R)
## [1] 10000 2
# modify column names for clear axes labeling
names(R) = c('x', 'y')
R$z = R$x^2 + R$y^2
head(R)
## x y z
## 1 -1.0000000 -1 2.000000
## 2 -0.9797980 -1 1.960004
## 3 -0.9595960 -1 1.920824
## 4 -0.9393939 -1 1.882461
## 5 -0.9191919 -1 1.844914
## 6 -0.8989899 -1 1.808183
ggplot(R, aes(x = x,y = y, z = z)) + stat_contour()
```


## Quantiles and Box Plots 
- The r-percentile of a numeric dataset is the point at which approximately r percent of the data lie underneath, and approximately 100r percent lie above1.
- AKA the 0.r quantile.

```{r}
# display 0 through 100 percentiles at 0.1 increments
# for the dataset containing 1,2,3,4.
quantile(c(1, 2, 3, 4), seq(0, 1, length.out = 11))
## 0% 10% 20% 30% 40% 50% 60% 70% 80%
## 1.0 1.3 1.6 1.9 2.2 2.5 2.8 3.1 3.4
## 90% 100%
## 3.7 4.0
```

- median or 50 percentile (2nd quartile): half the data lies underneath and half above 
- 25 percentile (1st quartile): above 25% of the data 
- 75 percentile (3rd quartile): above 75% of the data
- inter-quartile range (IQR): interval between the first and third quartiles
  - covers the central 50% of the data 

- The + operator below adds the box plot geometry, flips the x, y coordinates, and removes the y-axis label.

```{r mtcars}
ggplot(mpg, aes("",hwy)) +
geom_boxplot() +
coord_flip() +
scale_x_discrete("")
```
- plotting box plots side by side: 
- uses `coord_flip()` to flip the plots horizontally 
- reorders the factors of the class variable to sort the box plots in order of increasing highway mpg medians 

```{r mtcars}
ggplot(mpg, aes(reorder(class, -hwy, median), hwy)) +
geom_boxplot() +
coord_flip() +
scale_x_discrete("class")
```

## qq-Plots 
- quantile-quantile plots 
- good for comparing two datasets 
- essentially scatter plots of quantiles of datasets 
- (look to p337 for what plot shapes imply)

Example: histograms for three datasets sampled from three bell-shaped
Gaussian distributions N(0, 1), N(0, 1), and N(0, 2)

```{r}
D = data.frame(samples = c(rnorm(200, 1, 1),
rnorm(200, 0, 1),
rnorm(200, 0, 2)))
D$parameter[1:200] = 'N(1,1)';
D$parameter[201:400] = 'N(0,1)';
D$parameter[401:600] = 'N(0,2)';
qplot(samples,
facets = parameter~.,
geom = 'histogram',
data = D)
```

- and now the qq-plots of the datasets (y axis) vs. a sample
drawn from the N(0, 1) distribution (x axis).

```{r}
D = data.frame(samples = c(rnorm(200, 1, 1),
rnorm(200, 0, 1),
rnorm(200, 0, 2)));
D$parameter[1:200] = 'N(1,1)';
D$parameter[201:400] = 'N(0,1)';
D$parameter[401:600] = 'N(0,2)';
ggplot(D, aes(sample = samples)) +
stat_qq() +
facet_grid(.~parameter)
```


## Data Preparation

1. create a df with data and informatively named columns 
2. create a plot with `plot`, `qplot`, or `ggplot`

creating a df: `R = data.frame(name = vec1, ages = vec2, salary = vec3)`
changing column names: `names(R) = c("last.name", "age", "annual.income")`
`rbind` and `cbind` add additional rows or columns to a df 

creating a df example
```{r}
A = list(a = c(1, 2), b = c(3, 4), c = c(5, 6))
stack(A)
```

graphing the df:
```{r}
x_grid = seq(-8, 8, length.out = 100)
gaussian_function = function(x, s) exp(-x^2/(2*s^2))/(sqrt(2*pi)*s)
R = stack(list('sigma=1' = gaussian_function(x_grid, 1),
'sigma=2' = gaussian_function(x_grid, 2),
'sigma=3' = gaussian_function(x_grid, 3),
'sigma=4' = gaussian_function(x_grid, 4)))
names(R) = c("y", "sigma");
R$x = x_grid
head(R)
## y sigma x
## 1 5.052271e-15 sigma=1 -8.000000
## 2 1.816883e-14 sigma=1 -7.838384
## 3 6.365366e-14 sigma=1 -7.676768
## 4 2.172582e-13 sigma=1 -7.515152
## 5 7.224128e-13 sigma=1 -7.353535
## 6 2.340189e-12 sigma=1 -7.191919
qplot(x,
y,
color = sigma,
lty = sigma,
geom = "line",
data = R,
main = "Normal density for different sigma values",
xlab = "$x$",
ylab = "$f(x)$")
```





