---
title: "ch11-data-processing"
author: "jenngeorgevich"
date: "9/19/2017"
output: html_document
---
## Data Processing 

## Missing Data 
MCAR: missing completely at random 
  - the prob of an observation mission does not depend on observed or unobserved measurements
MAR: missing at random
  - given the observed data, the prob that data is missing does not depend on the unobserved data
  
Some ways to deal with missing data:
- remove rows containing missing data
- replace missing entries with a substitute variable
- estimate a probability model for the missing variable and replace the missing values with samples from the probability model 

## Outliers 
two views:
- corrupted values (human error)
- non-corrupt, but substantially unlikely values 

Robustness: data analysis procedures not sensitive to outliers
  -non-robust ex: computing the mean of n numbers 
  -robust: median of n numbers 
  
Techniques for dealing with outliers:
- Truncating: removing all outliers 
- Winsorization: Shrink outliers to the edge of the main data. 
  - example: replace outliers with the most extreme of accepted values
- Robustness: use a robust procedure to analyze the data 

How to determine outliers:
- below the alpha percentile or above the 100 - alpha percentile for some small alpha > 0 
- if we assume a symmetric distribution, this ^ corresponds to being more than c std deviations away from the mean. 
- compute the std deviation or percentiles after removing the most extreme values to avoid corrupting them with the outliers 

Example removing outliers
```{r}
# sample values from Gaussian distribution 
original_data = rnorm(20)
# create an outlier 
original_data[1] = 1000
# sort the data and exclude lowest and highest values 
filtered_data = original_data[3:18]
# find limits using std devs 
lower_limit = mean(filtered_data) - 5 * sd(filtered_data)
upper_limit = mean(filtered_data) + 5 * sd(filtered_data)
# exclude values 
not_outlier_ind = (lower_limit < original_data) &
(original_data < upper_limit)
```

Example windsorizing data using `windsorize` from `robustHD`

```{r}
library(robustHD)
original_data = c(1000, rnorm(10))
print(original_data)
## [1] 1000.0000000 -0.2925366 1.8952183
## [4] 0.2895131 0.2843515 -0.5394700
## [7] -0.2802351 -0.8678683 0.1259644
## [10] 0.6260071 -0.3043207
print(winsorize(original_data))
## [1] 1.4018458 -0.2925366 1.4018458
## [4] 0.2895131 0.2843515 -0.5394700
## [7] -0.2802351 -0.8678683 0.1259644
## [10] 0.6260071 -0.3043207
```

## Data Transformations 

### Skewness and Power Transformation
- sometimes data comes from a highly-skewed distribution
- it cant be described well by a common stat distribution 
- so use a transformation to map the data to a form that is well described by common distributions 

Power Transformations: transformations for non-negative values 
-see page 361 for equation
  - for lambda in the set of Real numbers 
  - fvlambda(x) = (x^lambda)/ lambda for lambda > 0 
  - fvlambda(x) = logx for lambda = 0, x> 0, lambda in set of Real numbers
  - fvlambda(x) = -(x^lambda -1)/ lambda for lambda < 0 
- it maps x to x^lambda 
- choose different lambda values to remove left or right skewness:
  - remove left skew: lambda > 1 
  - remove right skew: lambda < 1 
  - select by trying different values and looking at the resulting histograms, or by using the maximum likelihood method 
  
Diamonds Example: 

```{r}
diamondsSubset = diamonds[sample(dim(diamonds)[1], 1000),]
qplot(price, data = diamondsSubset)
```
transformed histogram: addresses the right skew 

```{r}
qplot(log(price), size = I(1), data = diamondsSubset)
```

transforming data shows a linear relationship on a log-log scale whereas the untransformed data didn't look like much. 

```{r}
qplot(carat,
price,
size = I(1),
data = diamondsSubset)

#transform!
qplot(carat,
log(price),
size = I(1),
data = diamondsSubset)

qplot(log(carat),
price,
size = I(1),
data = diamondsSubset)

qplot(log(carat),
log(price),
size = I(1),
data = diamondsSubset)
```

displaying the original variables on logarithmix axes for more readbility 
```{r}
qplot(carat,
price,
log = "xy",
size = I(1),
data = diamondsSubset)
```

Animals example 
```{r}
library(MASS)
#hard to see anything of interest 
qplot(brain, body, data = Animals)
#transform!
qplot(brain, body, log = "xy", data = Animals)
# transforming reveals some outliers 
```


### Binning 
- Binning (AKA discretization): divide a numeric variable's range into bins and replacing the variable with a number representing the corresponding bin 
  - like rounding 
- Binarization: a type of binning
  - replace variables with 0 or 1 

### Indicator Variables 
- "a technique that replaces a variable x (numeric, ordinal, or categorical) taking k values with a binary k-dimensional vector v, such that v[i] (or vi in mathematical notation) is one if and only if x takes on the i-value in its range. Thus, the variable is replaced by a vector that is all zeros, except
for one component that equals one corresponding to the variable value."
- sometimes used with binning: make k bins, then create a k dimensional indicator variable 
-useful for:
  - ordinal or categorical data 
  - capturing non-linear relationships by training a linear model on the result of: transforming data that doesn't follow a common distribution, binning it, and creating indicator vectors for the binned values
  - better scalability: easier to compute  because indicator variables are binary 

## Data Manipulation 

### Random Sampling, Partitioning, and Shuffling
- random sampling without replacement: on each selection, exclude selected rows from the group being sampled 
- random sampling with replacement: do not remove selected rows from the group being sampled from (may select the same row multiple times)

```{r}
D = array(data = seq(1, 20, length.out = 20), dim = c(4, 5))
print(D)
#get a sample of rows
sampled_row_indices = sample(1:4, 3, replace=FALSE)
# make a new array or df of the sampled rows 
D_sampled = D[sampled_row_indices,]
print(D_sampled)

```

random permutation of k objects: `sample(k, k)`

### Concatenations and Joins 
sometimes you need to make a single data frame out of multiple sources 
(book shows how to do this in python)

from http://www.statmethods.net/management/merging.html:
  - merge 2 data frames by ID: `total <- merge(data frameA,data frameB,by="ID")`
  - merge two data frames by ID and Country: `total <- merge(data frameA,data frameB,by=c("ID","Country"))`
  - add rows to a dataframe if the 2 data frames have the same column names:
  `total <- rbind(data frameA, data frameB)`

### Tall Data and Wide Data 
- tall format: array or df contains one or more columns that is a unique identifier and an additional column represents value 
  - may have many rows but few columns

example: 
2015/01/01 apples 200
2015/01/01 oranges 150
2015/01/02 apples 220
2015/01/02 oranges 130

- wide format: may represent info with multiple columns rather than multiple rows 
  - usually simpler to analuze 

example: 

Date apples oranges
--------------------------
2015/01/01 200 150
2015/01/02 220 130

### Reshaping Data 
- can use the reshape2 package in R 

example making a wide df tall
```{r}
library(reshape2)
head(smiths)
melt(smiths, id = c(1, 2))
```

- use `acast` (for arrays) or `dcast` (for dfs) to do the inverse of melt 
```{r}
head(tips)
tipsm = melt(tips,
id = c("sex","smoker","day","time","size"))
head(tipsm)
# mean is the aggregating function 
dcast(tipsm,
sex~variable,
fun.aggregate = mean)
```


### The Split-Apply-Combine Framework 
Process:
  - split the df to form smaller arrays or dfs 
  - apply some operation to each of the other smaller arrays or dfs 
  - combine the results into a single meaningful array or df 
  
- plyr R package (see p 384 for chart )

example 
```{r}
library(plyr)
head(baseball)
# count number of players recorded for each year
bbPerYear = ddply(baseball, "year", "nrow")
head(bbPerYear)
qplot(x = year,
y = nrow,
data = bbPerYear,
geom = "line",
ylab="number of player seasons")
```

example 2
```{r}
bbMod=ddply(baseball,
"year",
summarise,
mean.rbi = mean(rbi, na.rm = TRUE))
head(bbMod)
qplot(x = year,
y = mean.rbi,
data = bbMod,
geom = "line",
ylab = "mean RBI")
```
more examples in book
